---
title: "Detecting High-Quality Signals of Adverse Drug-Drug Interactions"
author: "Chen Zhan"
date: "15/04/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The page provides a demo of R codes of our ADDI signal detection algorthim, which proposed in  _Detecting High-Quality Signals of Adverse Drug-Drug Interactions_.

__Abstract:__

As a medicine safety issue, Drug-Drug Interaction (DDI) may become an unexpected threat for causing Adverse Drug Events (ADEs). There is a growing demand for computational methods to efficiently and effectively analyse large-scale data to detect signals of Adverse Drug-drug Interactions (ADDIs). In this paper, we aim to detect high-quality signals of ADDIs which are non-spurious and non-redundant. We propose a new method which employs the framework of Bayesian network to infer the direct associations between the target ADE and medicines, and uses domain knowledge to facilitate the learning of Bayesian network structures. To improve efficiency and avoid redundancy, we design a level-wise algorithm with pruning strategy to search for high-quality ADDI signals. We have applied the proposed method to the United States Food and Drug Administration's (FDA) Adverse Event Reporting System (FAERS) data. The result shows that 54.45% of detected signals are verified as known DDIs and 10.89% were evaluated as high-quality ADDI signals, demonstrating that the proposed method could be a promising tool for ADDI signal detection.

The high-quality signals detect by this experiment, as well as their evaluation results, are listed in http://nugget.unisa.edu.au/ADDI/Result_Rhab.xlsx

Step 1: Load the FDA's AERS data.
Its public version can be download from the FDA's website: https://fis.fda.gov/extensions/FPD-QDE-FAERS/FPD-QDE-FAERS.html

We also attached a copy of 2018Q3 dataset of FAERS on this server, which can be found at http://nugget.unisa.edu.au/ADDI/ascii/

```{r}
#use 2018 Q3 data as an example, the FAERS is already structured data

library(tictoc) # record processing time
tic('total')

tic('load FDA dataset' )
library(readr) 

#DRUG dataset contains information of medicines mentioned in reports
DRUG18Q3 <- read_delim("~/rds/FDA/ascii/DRUG18Q3.txt", 
                       "$", escape_double = FALSE, trim_ws = TRUE)

#REAC dataset contains adverse event information mentioned in reports
REAC18Q3 <- read_delim("~/rds/FDA/ascii/REAC18Q3.txt", 
                       "$", escape_double = FALSE, trim_ws = TRUE)

#DEMO dataset demographic informtion of patients mentioned in reports
DEMO18Q3 <- read_delim("~/rds/FDA/ascii/DEMO18Q3.txt", 
                       "$", escape_double = FALSE, trim_ws = TRUE)


head(DEMO18Q3)
head(DRUG18Q3)
head(REAC18Q3)
toc()
```

Data cleaning and preprocessing, including drug name standardisation and records de-duplication/duplication checking

The drug name standardisation is recommended using MedEx, a natural language processing tool that designed for the extraction of medication information from clinical notes, which can be found in https://sbmi.uth.edu/ccb/resources/medex.htm.
Here we present a simplified solution for this process.

The idea of a simplified solution is to impute the missing generic names by checking other records, and to map a drug to his multiple generic names (if it has) by splitting and matching texts.

The criterion of duplication checking, please refer our paper.

```{r}

tic('data cleaning and reshaping')

ADR_of_interest = 'Rhabdomyolysis'
N = length(unique(REAC18Q3$primaryid))


length(DEMO18Q3$primaryid)
length(unique(DEMO18Q3$caseid)) 
print('There is no duplicated records if above two values are equal')

pid_list = unique(REAC18Q3[REAC18Q3$pt == ADR_of_interest,]$primaryid)
cand_drugs_raw = unique(DRUG18Q3[DRUG18Q3$primaryid %in% pid_list,]$prod_ai)


cand_drugs = lapply(strsplit(cand_drugs_raw,'\ |[\\]+'), unique)
cand_drugs_first = na.omit(unique(sapply(cand_drugs, function(x){
  x[[1]]
})))
cand_drugs_first = setdiff(cand_drugs_first, c("RED")) # manually rm wrong vocabulary
test = (lapply(cand_drugs_first, function(x){
  grep(x,cand_drugs_raw,ignore.case = T)
}))
names(test) = cand_drugs_first

#test is a list whose element's name is standardised drug name and its value is index of cand_drugs_raw

```

Reshape the data to a sparse matrix, which rows are reports, columns are drugs and its values are the indicator of whether drugs were mentioned in the reports.

Concate the outcome with data, the outcome is the occurence of ADE of interest with each reports.

```{r}

data = as.data.frame(matrix(nrow = N, ncol = length(test), dimnames = list(unique(REAC18Q3$primaryid), names(test))))
for (i in 1:length(test)){
  data[as.character(unique(DRUG18Q3[DRUG18Q3$prod_ai %in% cand_drugs_raw[test[[i]]],]$primaryid)), names(test[i])] = 1
  #print(i)
}
data[is.na(data)] = 0

Outcome = matrix(nrow = N,  ncol = 1, dimnames = list(rownames(data), ADR_of_interest) )
Outcome[as.character(unique(REAC18Q3[REAC18Q3$pt == ADR_of_interest,]$primaryid)),] = 1
Outcome[is.na(Outcome)] = 0

new_data = as.data.frame(cbind(data, Outcome))

toc()

```

Retrieve and organise domain knowledge from SIDER, which can be download from http://sideeffects.embl.de/download/

In this demo, we manually retrieve it from website for the case study.

```{r}

t = '
Benicar-HCT
EACA
K779
LY146032
Zyprexa Relprevv
abacavir
abacavir-lamivudine
aliskiren
aminophylline
amiodarone
amphetamine
amphotericin B
amprenavir
aripiprazole
asenapine
atorvastatin
benzathine penicillin
bezafibrate
bortezomib
bupropion
cabozantinib
candesartan
candesartan cilexetil
cefdinir
cerivastatin
ciprofloxacin
citalopram
clarithromycin
clomipramine
clozapine
colchicine
cyclophosphamide
cytarabine
darunavir
dasatinib
delavirdine
desflurane
didanosine
duloxetine
efavirenz
emtricitabine
entacapone
eprosartan
erlotinib
etravirine
everolimus
ezetimibe
famotidine
febuxostat
felbamate
fenofibrate
fenofibric acid
fluvastatin
fluvoxamine
fosamprenavir
foscarnet
fusidic acid
gabapentin
ganciclovir
gemfibrozil
hydrochlorothiazide
ifosfamide
imatinib
indapamide
indinavir
irbesartan
irbesartan-hydrochlorothiazide
lamivudine
lamotrigine
lenalidomide
losartan
lovastatin
lurasidone
maraviroc
melphalan
milnacipran
morphine
nefazodone
nelarabine
nelfinavir
nevirapine
niacin
ofloxacin
olanzapine
olmesartan
olmesartan medoxomil
oxaliplatin
paliperidone
pantoprazole
penicillin
pentamidine
pitavastatin
pramipexole
pravastatin
pregabalin
propofol
quetiapine
rabeprazole
raltegravir
retinoic acid
ribavirin
risperidone
rosuvastatin
saquinavir
sevoflurane
simvastatin
sorafenib
sparfloxacin
succinylcholine
sulfamethoxazole
sulfasalazine
sunitinib
tacrolimus
telmisartan
temsirolimus
tenofovir
tenofovir disoproxil fumarate
terbinafine
theophylline
thymidine
tipranavir
tizanidine
tolcapone
tolvaptan
trabectedin
trametinib
trien
trospium chloride
v 1784
valsartan
venlafaxine
ziconotide
zidovudine
zidovudine/lamivudine
ziprasidone
zonisamide
'

cit00 = read.table(text = t, sep = '\n', stringsAsFactors =F)[,1]

#preliminary check whether these medicines are in the analysis 
cit0 = toupper(cit00[sapply(cit00, function(x){
  sum(grepl(x, names(test), ignore.case = T))
})>0])
cit0 = cit0[cit0 %in% names(test)]

toc()

```

Level 1 (The codes beyong this line would be running, demostration purpose only)

alpha set as 0.01 for a demo purpose

How adaptively determine alpha, please refer our paper

```{r eval=FALSE}

library(bnlearn)

tic('level 1')

library(bnlearn)
ci_res = lapply(names(test), function(x){
  ci.test(x = x, y = colnames(Outcome), data = new_data)
})

ci_rank = matrix(unlist(lapply(ci_res, function(x){
  c(unname(x$statistic), x$p.value)
})), ncol = 2, byrow = T)
ci_rank = as.data.frame(ci_rank)
rownames(ci_rank) = names(test)
colnames(ci_rank) = c('x2', 'p')

ci_res_1 = lapply(setdiff(rownames(ci_rank[ci_rank[,2]<0.01,]), cit0), function(x){
  max(sapply(cit0, function(x,y){
    ci.test(x = x, y = colnames(Outcome),z = y, data = new_data)$p.value
  },x = x))
})

cit1 = setdiff(rownames(ci_rank[ci_rank[,2]<0.01,]), cit0)[ci_res_1<0.01]

toc()

```

Level 2

We require at least 4 cases for including a candidate combination into the test.

```{r eval=FALSE}

tic('level2')

rest_drugs = setdiff(setdiff(names(test), cit0), cit1)

final_res = list()
for (i in 1:length(rest_drugs)){
  for (j in (i+1):length(rest_drugs)){
    combined = data[,rest_drugs[i]] * data[,rest_drugs[j]]
    if (sum(combined)>4){# minimum 4 cases as support
      combined = matrix(combined, ncol = 1, byrow = T,dimnames = list(rownames(data), 'combined') )
      combined = as.factor(combined)
      temp = max(sapply(union(cit0,cit1), function(y){
        ci.test(x = combined, y = new_data[,colnames(Outcome)],z = new_data[,y])$p.value
      }))
      if (temp <0.1){ # only record results with p-value < 0.1, to save memory space
        final_res[length(final_res)+1] = temp
        names(final_res)[length(final_res)] =paste(rest_drugs[i], rest_drugs[j], sep = ' + ')
      }
    }
    #print(c(i,j))
  }
}
toc()

print(sort(final_res))

toc()

```